Mamary Gland Development Exploratory Data Analysis

### 2017-01-16 src/QCAnalysis.Rmd
QC As of now

## Cell Viability
```{r}
topFreq20 <- filter(fD, symbol %in% plotData[1:20,1]) %>% .$id
pD$prcntOfTop20 <- colSums(m[topFreq20,]!=0)/length(topFreq20)
mMito <- m[fD$Mitochondrial,]
pD$prcntMito <- colSums(mMito)/colSums(m)
cellViability <- ggplot(pD, aes(x=prcntMito, y=GenesDetected, color=Condition, shape=Replicate))+
    geom_point() +
    theme_bw()
cellViability
```
```{r}
smry <- pD %>%
    summarize(medianPrcntMito=median(prcntMito),
	      madPrcntMito=mad(prcntMito),
	      mGenesDetected=median(GenesDetected),
	      madGenesDetected=mad(GenesDetected),
	      mUmiSums=median(UmiSums),
	      madUmiSums=mad(UmiSums))
smryByGroup <- group_by(pD,Condition) %>%
    summarize(medianPrcntMito=median(prcntMito),
	      madPrcntMito=mad(prcntMito),
	      mGenesDetected=median(GenesDetected),
	      madGenesDetected=mad(GenesDetected),
	      mUmiSums=median(UmiSums),
	      madUmiSums=mad(UmiSums))
```

## QC Thresholding
```{r}
#simple thresholds
MitoCutOff <- 0.05
genesDetectedCutOff <- 500
LibSizeCutOff <- 500
```
Four different thresholds were applied:

1. Percentage of UMIs in Mitochondrial genes less than `r MitoCutOff`
2. A minimum of `r genesDetectedCutOff` genes detected per cell
3. A minimum of `r LibSizeCutOff` total UMIs per per cell
4. Not more than four standard deviations away from the #Genes~#UMIcount trend away (see below)
```{r}
#remove outliers from the genes detected per UMI trend 
fit <- loess(formula=log10(GenesDetected)~log10(UmiSums),data=pD)
pD$distFromDetectionTrend <- fit$residuals
upperL <- mean(fit$residuals)+4*sd(fit$residuals)
lowerL <- mean(fit$residuals)-4*sd(fit$residuals)
hist(fit$residuals, breaks=50, main="Histogramm of loess residuals \n and thresholds",
     xlab="Residuals")
abline(v=upperL,col="red")
abline(v=lowerL,col="red")

#Thresholding 
pD <- mutate(pD,
		  PassViability = prcntMito < MitoCutOff,
		  PassGenesDet = GenesDetected > genesDetectedCutOff,
		  PassLibSize = UmiSums > LibSizeCutOff,
		  PassDetTrend = distFromDetectionTrend < upperL & 
		      distFromDetectionTrend > lowerL,
		  PassAll= PassViability & PassGenesDet &
		      PassLibSize & PassDetTrend)
#Update QC plots
gdLibSize <- gdLibSize %+% pD 
gdLibSize <- gdLibSize + aes(color=PassDetTrend,shape=NULL) + 
    annotate("rect",xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=log10(genesDetectedCutOff), fill="grey", alpha=0.3) + 
    annotate("rect",ymin=-Inf, ymax=Inf, xmin=-Inf, xmax=log10(LibSizeCutOff), fill="grey", alpha=0.3) 
gdLibSize
cellViability <- cellViability %+% pD
cellViability <- cellViability + aes(color=PassViability) +
    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff
	     ,fill="grey", alpha=0.3) 
cellViability

kable(t(as.matrix(table(pD$Condition,pD$PassAll))),caption="Cells passing QC in different Conditions")
#Apply Thresholds
pD.filtered <- filter(pD, PassAll)
m.filtered <- m[,as.character(pD.filtered$barcode)]

#Thresholding on lowly expressed genes
expThreshold <- 0.1
keep <- rowMeans(m.filtered) > expThreshold
m.filtered <- m.filtered[keep,]
fD.filtered <- fD[keep,]
#Final matrix
dim(m.filtered)
```

## Normalization and first visualization
```{r}
clusters <- quickCluster(m.filtered)
pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters,positive=TRUE)
m.norm <- t(t(m.filtered)/pD.filtered$sf)
m.norm <- m.norm[,!is.na(colSums(m.norm))]
pD.filtered <- filter(pD.filtered, barcode %in% colnames(m.norm))
# 
fD.filtered <- mutate(fD.filtered, meanExpression=rowMeans(m.norm),
	     CV2=apply(m.norm,1,cv2),
	     dm=DM(meanExpression,CV2))

highVar <- dplyr::arrange(fD.filtered,desc(dm)) %>% .$id
topX <- 0.1*length(highVar)
fD.filtered$highVar <- fD.filtered$id %in% highVar[1:topX]
VarPlot <- ggplot(fD.filtered, aes(x=log10(meanExpression),y=log10(CV2), color=highVar)) +
    geom_point() +
    theme_bw()
VarPlot

@@

pD.filtered$PC1 <- pc$x[,1]
pD.filtered$PC2 <- pc$x[,2]
pD.filtered$tSNE1 <- tsn$Y[,1]
pD.filtered$tSNE2 <- tsn$Y[,2]

Acta2 <- filter(fD.filtered, symbol=="Acta2") %>% .$id
pD.filtered$Acta2 <- t(m.norm)[,Acta2]
# 
pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    theme_bw() 
pcaPlot

@@

tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Acta2+1), shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    theme_bw()
tsnePlot
```
## Save Data
```{r}
out <- list()
out[[1]] <- m
out[[2]] <- pD
out[[3]] <- fD
saveRDS(out,file="../data/Robjects/ExpressionList_QC.rds")
```

>>>>

## Cell Viability
```{r}
mMito <- m[fD$Mitochondrial,]
pD$prcntMito <- colSums(mMito)/colSums(m)
cellViability <- ggplot(pD, aes(x=prcntMito, y=GenesDetected, color=Condition, shape=Replicate))+
    geom_point() +
    theme_bw()
cellViability
```
```{r}
smryByGroup <- group_by(pD,Condition) %>%
    summarize(medianPrcntMito=median(prcntMito),
	      madPrcntMito=mad(prcntMito),
	      threshold_PrcntMito=0.05,
	      mGenesDetected=median(log10(GenesDetected)),
	      madGenesDetected=mad(log10(GenesDetected)),
	      threshold_GenesDetected=max(mGenesDetected-3*madGenesDetected,log10(500)),
	      mUmiSums=median(log10(UmiSums)),
	      madUmiSums=mad(log10(UmiSums)),
	      threshold_UmiSums=max(mUmiSums-3*madUmiSums,log10(900))) %>%
    select(Condition,starts_with("threshold")) 
kable(smryByGroup)
```

## QC Thresholding
```{r}
#simple thresholds
MitoCutOff <- 0.05
```
```{r}
#Thresholding per group
#inititae Df
pD <- mutate(pD,
	     ThresholdViability = 0,
	     ThresholdGenesDet = 0,
	     ThresholdLibSize = 0)

grps <- as.character(unique(pD$Condition))
for (grp in grps) {
    thrs <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% as.vector()
    names(thrs) <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% rownames()
    pD <- mutate(pD,
		 ThresholdViability= ifelse(Condition==grp, MitoCutOff, ThresholdViability),
		 ThresholdGenesDet= ifelse(Condition==grp, 10^thrs["threshold_GenesDetected"],ThresholdGenesDet),
		 ThresholdLibSize= ifelse(Condition==grp, 10^thrs["threshold_UmiSums"],ThresholdLibSize))
}

pD <- mutate(pD,
	     PassViability=prcntMito < ThresholdViability,
	     PassGenesDet=GenesDetected > ThresholdGenesDet,
	     PassLibSize=UmiSums > ThresholdLibSize,
	     PassAll= PassViability & PassGenesDet & PassLibSize)

gdHist <- ggplot(pD, aes(x=log10(GenesDetected),y=..density..)) +
    geom_histogram(fill="white",color="black") +
    geom_vline(data=smryByGroup,aes(xintercept=threshold_GenesDetected),color="red",lty="longdash") +
    facet_wrap(~Condition) +
    theme_bw()
gdHist

libSizeHist <- ggplot(pD, aes(x=log10(UmiSums),y=..density..)) +
    geom_histogram(fill="white",color="black") +
    geom_vline(data=smryByGroup,aes(xintercept=threshold_UmiSums),color="red",lty="longdash") +
    facet_wrap(~Condition) +
    theme_bw()
libSizeHist

cellViability <- cellViability %+% pD
cellViability <- cellViability + aes(color=PassViability) +
    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff
	     ,fill="grey", alpha=0.3) 
cellViability

kable(t(as.matrix(table(pD$Condition,pD$PassGenesDet))),caption="Cells passing Gene Detection QC")
kable(t(as.matrix(table(pD$Condition,pD$PassLibSize))),caption="Cells passing Library Size QC")
kable(t(as.matrix(table(pD$Condition,pD$PassViability))),caption="Cells passing Viability QC")
kable(t(as.matrix(table(pD$Condition,pD$PassAll))),caption="Cells passing all QC requirements")
#Apply Thresholds
pD.filtered <- filter(pD, PassAll)
m.filtered <- m[,as.character(pD.filtered$barcode)]

#Thresholding on lowly expressed genes
isexpThreshold <- 10
expThreshold <- 100*isexpThreshold
keep1 <- rowSums(m.filtered!=0) > isexpThreshold
keep2 <- rowSums(m.filtered) > expThreshold
fD$keep <- keep1 & keep2
m.filtered <- m.filtered[fD$keep,]
fD.filtered <- fD[fD$keep,]
#Final matrix
dim(m.filtered)

qcPCA <- select(pD.filtered, GenesDetected, UmiSums, prcntMito) %>%
    mutate(GenesDetected=log10(GenesDetected),
	   UmiSums=log10(UmiSums))
qcpc <- prcomp(qcPCA)
qcpD <- pD.filtered %>%
    mutate(PC1=qcpc$x[,1],
	   PC2=qcpc$x[,2])
qcpcaPlot <- ggplot(qcpD, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    ggtitle("PCA plot on QC stats after QC") +
    theme_bw() 
qcpcaPlot

```

## Normalization and first visualization
```{r}
clusters <- quickCluster(m.filtered)
pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters,positive=TRUE)
plot(log10(colSums(m.filtered))~log10(pD.filtered$sf),main="Library Size versus Size Factors")
m.norm <- t(t(m.filtered)/pD.filtered$sf)
m.norm <- m.norm[,!is.na(colSums(m.norm))]
pD.filtered <- filter(pD.filtered, barcode %in% colnames(m.norm))
# 
fD.filtered <- mutate(fD.filtered, meanExpression=rowMeans(m.norm),
	     CV2=apply(m.norm,1,cv2),
	     dm=DM(meanExpression,CV2))

highVar <- dplyr::arrange(fD.filtered,desc(dm)) %>% .$id
topX <- floor(0.2*length(highVar))

fD.filtered$highVar <- fD.filtered$id %in% highVar[1:topX]
VarPlot <- ggplot(fD.filtered, aes(x=log10(meanExpression),y=log10(CV2), color=highVar)) +
    geom_point() +
    theme_bw()
VarPlot

@@

pD.filtered$PC1 <- pc$x[,1]
pD.filtered$PC2 <- pc$x[,2]
pD.filtered$PC3 <- pc$x[,3]
pD.filtered$tSNE1 <- tsn$Y[,1]
pD.filtered$tSNE2 <- tsn$Y[,2]

Acta2 <- filter(fD.filtered, symbol=="Acta2") %>% .$id
pD.filtered$Acta2 <- t(m.norm)[,Acta2]
Fcer1g <- filter(fD.filtered, symbol=="Fcer1g") %>% .$id
pD.filtered$Fcer1g <- t(m.norm)[,Fcer1g]
# 
pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    theme_bw() 
pcaPlot

@@

tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Acta2+1), shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    theme_bw()
tsnePlot

```
***
## Heatmap on top50
```{r, fig.height=10}
gns <- highVar[1:50]
m.ht <- log2(m.filtered[gns,]+1)
rownames(fD) <- fD$id
rownames(m.ht) <- fD[rownames(m.ht),"symbol"]
colanno <- data.frame("Condition"=pD.filtered$Condition)
rownames(colanno) <- as.character(pD.filtered$barcode)
library(pheatmap)
pheatmap(m.ht, show_colnames=FALSE, annotation_col=colanno, show_rownames=TRUE)

```
## Save Data
```{r}
out <- list()
out[[1]] <- m
out[[2]] <- pD
out[[3]] <- fD
saveRDS(out,file="../data/Robjects/ExpressionList_QC.rds")
```
